Blaming the Typeless: \\ Scalable, Human-Centric Python Localization

### 2018-08-21 paper/main.tex
approach section

@@

The relative costs and benefits of static vs. dynamic error detection are
well-studied. Despite this, however, many currently-available techniques
are ill-suited to address the problem of localization for type errors in
dynamic languages. First, relying on the extant language interpreter
requires test inputs~\cite{FIXME}, may involve error messages that are
difficult for novices to understand~\cite{FIXME}, and may misleadingly
implicate the symptom rather than the cause~\cite{FIXME}.  Second, a
significant amount of development is carried out via IDE or web interfaces,
which require tools that operate on code fragments~\cite{Guo2013-vu}.
Third, attempts to retrofit static type systems into dynamic languages,
such as TypeScript for JavaScript~\cite{FIXME} and FIXME for
Python~\cite{FIXME} are promising but not yet as widely
deployed~\cite{FIXME}. Finally, research tools and algorithms for fault
localization either require rich static types (e.g., as in
Mycroft~\cite{FIXME} or Sherrloc~\cite{FIXME}), require test cases (e.g.,
as in Tarantula~\cite{FIXME}), or produce voluminous ranked output lists
that have been found to be unhelpful in general and less useful to novices
in particular~\cite[Sec.~5.1]{orso-parnin}.

@@

\section{Motivating Example}

@@

\section{Overview and Approach}

We present an algorithm for accurately localizing faults~\cite{tarantula} in
dynamically-typed Python programs that exhibit non-trivial uncaught runtime
exceptions. We do not consider syntax errors or references to undefined
variables. Our algorithm uses machine learning models based on static and
dynamic features to implicate suspicious expressions. Since studies have
found that voluminous fault localization output is not useful to
developers~\cite{orso-parnin,orso-parnin2015}, we focus on producing
Top-1 and Top-3 rankings.

@@

\subsection{Feature Extraction}
\label{sec-features}

Our key insight is that FIXME-INSIGHT. Are a result, we make use of
static, dynamic and contextual information. Static features, such as FIXME,
are effective at capturing FIXME-INFORMATION. By contrast, dynamic
features, such as FIXME, support the modeling of FIXME. Finally, we also
include contextual information, such as FIXME, which allows our algorithm
to accurately handle FIXME. We consider the empirical justification of
these features in Section~\ref{sec-eval}.

We calculate features for each statement and expression in the program.

\begin{itemize}

\item \emph{Static / syntactic features}
\begin{enumerate}
    \item What kind of statement/expression it is, e.g. assignment / return
     / import (for statements) or variable / literal / application (expressions).
    \item Size (the number of AST nodes in the subtree rooted at this node)
\end{enumerate}

\item \emph{Dynamic features}
We run each program through an instrumented interpreter (i.e. a modified PythonTutor
backend, which itself is based on the BDB debugging library). This lets us
compute the following features:
\begin{enumerate}
    \item What type(s) does the expression have at runtime? (The values of this feature
    can be a type like int, or the special values "unknown" if the expression is
    never successfully evaluated and "multiple" if it changes type.)
    \item Is it part of the error slice?
    \item Is it the node at which the program actually crashes?
\end{enumerate}
(Since this interpreter works at the granularity of lines and we want to work with
expressions as well, we first convert each program to ANF.)

We compute the error slice by running the program and building a dependency
graph where node A depends on node B if either B defines a variable that A uses,
or B affects control flow allowing A to run.

We also check what uncaught exception is thrown by the program (TypeError,
KeyError, etc), and add that as a constant feature to all vectors derived from that program.

\item \emph{Contextual features} After all other features are computed so we have a
vector $v_e$ representing expression $e$, we set $v_e^{+context} = v_e \circ
v_p \circ v_{c_1} \circ v_{c_2} \circ v_{c_3}$ where $p$ is the expression's parent
in the AST and $c_i$ are its children. This allows us to recover some of the
program structure that would otherwise be completely lost when we convert the
structured program into an unstructured bag of feature vectors.

\item \emph{Label} Each feature vector is labeled by whether that expression/statement
changed between the crashing and fixed versions. (This is our proxy for whether
that node is to blame for the crash).

\end{itemize}

Each vector is then one-hot encoded (e.g. feature "Type : Int" becomes set
of features "Type-Int : 1", "Type-Bool : 0", "Type-Str : 0", etc.) for use with
standard machine learning tools.

\subsection{Machine Learning and Modeling}
\label{sec-model}

The models we train (inc. decision trees/forests,
MLP, GBM) can then rank expressions of a test program in order of how likely they
are to be the source of the bug, so we can compute Top-k accuracies: how often
is the true program change located in the model's first k outputs.

\section{Approach, in more detail}

We follow the overall strategy of \cite{learning-to-blame} and represent each
buggy program by a Bag of Abstracted Terms. In our setting, a `term' means either
a statment or an expression.

For each term in the program we compute static, dynamic,
and contextual features. While it is easy to read some features off of the surface
syntax, we expect that some features which are only accessible dynamically will also
be useful, for example the type of each expression. Since programs are highly
structured and expressions gain meaning in relation to their surroundings,
we also expect contextual features to be useful.

\subsection{Static features}

\emph{Syntactic form} We expect certain kinds of terms to be more prone to bugs
than others - for example, students might have more trouble with loop conditions
than with simple assignments. Thus the first feature we consider is what kind of node the term
is in the AST. For statements this has values like Return or Import; for expressions
Variable or Application.

\emph{Size} We also count the number of descendent nodes in the subtree rooted at that node.

\subsection{Dynamic features}

We first convert each buggy program to an equivalent program that is closer
to ANF, then run it through the PythonTutor backend. This backend is based on the
Python debigging library BDB; it returns to us a trace of the program along with
the state of the heap at each execution step. We use this information to compute
the following features for each term:

\emph{Type} An expression's type should be a very useful feature. Not only are
incompatible types a sign of an impending crash, but some types may be inherently
suspect whenever they appear, e.g. there are very few reasons to have a variable
of type NoneType. Our possible values for the Type feature include all the basic Python
types (such as int and tuple) as well as a few special values:
\begin{itemize}
    \item IsStatement - Given to all statements, since our terms include both
    statments and expressions but only expressions actually have types.
    \item Unknown - Given to expressions that are never evaluated in the dynamic
    trace.
    \item Multiple - Given to expressions which are evaluated multiple times in
    a trace and do not always have the same type.
\end{itemize}

\emph{Slice}
\begin{figure}
\begin{lstlisting}
x = input() # The user inputs 0
if x != 0:
  x = 1
print("One divided by your number is: %d" % (1 / x))
\end{lstlisting}
\caption{In this program, the programmer may have intended the placeholder
  assignment \texttt{x = 1} to be used in case the user input 0, and used the
  incorrect condition \texttt{x != 0}, rather than \texttt{x == 0}. However, the
  condition expression will not be included in the slice (which would include
  lines 1 and 4), because the presence of
  the conditional does not affect the behavior of the program.
}
\label{slice-downside-example}
\end{figure}

\begin{figure}
\begin{lstlisting}
while true:
  x = input()
  if x != 0:
    break
  print("One divided by your number is: %d" % (1 / x))
\end{lstlisting}
\caption{
  This program's mistake is the same as before,
  but instead they try to handle the
  bad input by breaking out of the input loop. When computing the slice, our
  method would record a control dependency between the condition on line 3 and
  the print statement on line 5, placing the bad condition in the program slice.
}
\label{early-break}
\end{figure}

The goal of this feature, roughly equivalent in purpose to the type error slice in
\cite {learning-to-blame}, is to help eliminate terms that cannot be the source of
the crash. We compute a dynamic program slice \cite{KOREL1988155}, the set of
terms that contributed at runtime to the observed exception, and turn this
feature on for terms in the slice. We follow the basic approach of
\cite{KOREL1988155, KOREL1990187}, building a graph of data and control
dependencies. We then traverse the graph backwards, starting at the execution
step where the error occurred, to collect the set of terms that the excepting
line transitively depended on. This excludes lines that could not have caused the
exception, such as lines that never ran, or lines that ran but had no connection
to the step where the exception happened.

A downside of using the dynamic slice, in the case of control dependencies, is
that it excludes terms whose presence does not affect whether the exception
happened, but which could have had an effect if the term was different. In
\figref{slice-downside-example}, the conditional statement is never run, so the
removal of the conditional would not have any effect on the program's behavior,
and it would not be included in the program slice. The dynamic slice misses the
incorrect test condition because the mistake itself caused the conditional
statement and its body to become irrelevant.

To overcome this issues in the ``early return'' or ``early break'' case
(\figref{early-break}), we check if a break, return, or other statement for
breaking out of structured control flow is present inside of a conditional
statement. We then add dependencies in the execution trace's dependency graph
between the enclosing conditional and the statements that would have been skipped
by the break or return. This is insufficient to overcome all related problems,
however, so unlike in prior work, we do not treat membership in the slice
as a hard constraint.

\emph{Crash location} As discussed later in the evaluation, the precise term that
causes Python to crash is in fact the one the user decides to change about 30\% of
the time, so we use it as a feature (and expect it to be a high-signal one, too.)

\emph{Exception type} The type of error thrown by the program is useful context for
finding the error. For example, if asked how likely a random `0` is of being a bug, you'd give
a higher answer if you knew that the program eventually crashes with a division by
zero exception. Thus we attach the exception type as a feature to \emph{every}
vector derived from that program.

\subsection{Contextual features}

The approach of representing a program by a Bag of Abstracted Terms, if implemented
naively, runs into the problem that the terms lose their context in the program
as a whole. For example, the features for the term `0` include its type (int) and
syntactic description (int literal), but there is no way that's enough to decide
if it's innocuous, like the 0 in `x = 0`, or clearly the source of a bug, like the
0 in `x / 0`. To restore some of this context, we introduce contextual features.
This involves duplicating all our other features four times - once for the parent
and once each for a term's first three children in the AST. So since `x / 0` has
the feature SyntacticForm = DivisionOp, the `0` of `x / 0` has the feature
ParentSyntacticForm = DivisionOp. Similarly, since `0` has SyntacticForm = IntLiteral,
`x / 0` has Child1SyntacticForm = IntLiteral. In order to make all feature vectors
the same length, nodes that have no parent or fewer than three children have the
relevant features filled out with a special NotApplicable value.

\subsection{Machine learning}

>>>>

\newcommand{\fixme}[1]{\textcolor{red}{FIXME: #1}}

@@

The relative costs and benefits of static vs. dynamic error detection are
well-studied. Despite this, however, many currently-available techniques
are ill-suited to address the problem of localization for type errors in
dynamic languages. First, relying on the extant language interpreter
requires test inputs~\cite{FIXME}, may involve error messages that are
difficult for novices to understand~\cite{FIXME}, and may misleadingly
implicate the symptom rather than the cause~\cite{FIXME}.  Second, a
significant amount of development is carried out via IDE or web interfaces,
which require tools that operate on code fragments~\cite{Guo2013-vu}.
Third, attempts to retrofit static type systems into dynamic languages,
such as TypeScript for JavaScript~\cite{FIXME} and FIXME for
Python~\cite{FIXME} are promising but not yet as widely
deployed~\cite{FIXME}. Finally, research tools and algorithms for fault
localization either require rich static types (e.g., as in
Mycroft~\cite{FIXME}, Nate~\cite{FIXME} or Sherrloc~\cite{FIXME}), require
test cases (e.g., as in Tarantula~\cite{FIXME}), or produce voluminous
ranked output lists that have been found to be unhelpful in general and
less useful to novices in particular~\cite[Sec.~5.1]{orso-parnin}.

@@

\section{Motivating Example}
\label{sec-motex}

@@

\section{Algorithm Overview}

We present an algorithm for accurately localizing faults~\cite{tarantula} in
dynamically-typed, beginner-written Python programs that exhibit
non-trivial uncaught runtime exceptions. We do not consider syntax errors
or references to undefined variables. Our algorithm uses machine learning
models based on static, dynamic, contextual and slicing features to
implicate suspicious expressions. Since studies have found that voluminous
fault localization output is not useful to
developers~\cite{orso-parnin,orso-parnin2015}, we focus on producing Top-1
and Top-3 rankings.

@@

Drawing inspiration from localization algorithms such as
Nate~\cite{learning-to-blame} and the natural language processing term
frequency vector (or ``bag of words'') model~\cite{FIXME}, we represent
each buggy program as a ``bag of abstracted terms''. A \emph{term} is
either a statement of expression.

\subsection{Model Feature Intuition and Extraction}

We observe that many type errors admit multiple logically valid resolutions
(see \secref{motex}): we thus cannot expect to match programmer desires
through type constraints alone, and instead propose to leverage information
from multiple modalities. We use static features to capture structured
program meaning, contextual features to capture the relationship between a 
program fragment and its environment, and dynamic features (including slice
information) to reason about conditional behavior.

For each term in the program we compute static, dynamic, and contextual
features. While some features can be computed directly from syntactic
information, others are only only accessible dynamically. In a
dynamically-typed language such as Python, this includes types of each
expression. Since programs are highly structured and expressions gain
meaning in relation to their surroundings, we also include contextual
features.

While static features can be computed via parsing, dynamic features require
executing the program fragment. To calculate dynamic features, we first
convert each buggy program to an equivalent program that is closer to ANF.
\fixme{WRW has no idea what ``ANF'' means in this context, and this is the
first and only time it appears in the paper text. Someone else should
rewrite this.} We define ANF to be FIXME and ``closer'' to mean FIXME in
this context. We then run the resulting program through the PythonTutor
backend~\cite{FIXME}.  This backend is based on the Python debigging
library BDB~\cite{FIXME}; it returns a trace of the program along with the
state of the heap at each execution step. That trace is used to compute
dynamic features.

\subsection{Model Features}
\label{sec-features}

\subsubsection{Syntactic Forms (Static)}

We hypothesize that certain syntactic categories of terms may be more
prone to bugs than others, especially for beginner programmers.  For
example, students might have more trouble with loop conditions than with
simple assignments. The first feature we consider is the syntactic
category of node. This feature is categorical, using standard abstract
syntax tree categories such as Return or Import for statements and Variable 
or Application for expressions.

\subsubsection{Expression Size (Static)}

This numeric feature counts the number of descendent nodes in the subtree
rooted at that node. Our intuition is that larger, more complex expressions
may be more prone to faults.

\subsubsection{Type (Dynamic)}

Although type constraints alone cannot always match human intuition for 
fault localization problems, types still contain very useful information.
We observe that some types may be inherently suspect whenever they appear,
especially for beginner-written code. For example, there are very few
reasons to have a variable of type \lt{NoneType} in Python. We model
type information as a categorical feature. Possible
values include all the basic Python types (such as \lt{int}
and \lt{tuple}) as well as three special values:
\begin{itemize}
    \item \lt{Statement} --- Given to all statements, since our terms
    include both statments and expressions but only expressions 
    have types
    \item \lt{Unknown} --- Given to expressions that are never evaluated in the dynamic
    trace
    \item \lt{Multiple} --- Given to expressions which are evaluated multiple times in
    a trace and do not always have the same type
\end{itemize}

\subsubsection{Slice (Dynamic)}
\begin{figure}
\begin{lstlisting}
x = input() # The user inputs 0
if x != 0:
  x = 1
print("One over your number is: %d" % (1 / x))
\end{lstlisting}
\caption{In this program, the programmer may have intended the placeholder
  assignment \lt{x = 1} to be used in case the user input 0, and used the
  incorrect condition \lt{x != 0}, rather than \lt{x == 0}. However, the
  condition expression will not be included in the slice (which would include
  lines 1 and 4), because the presence of
  the conditional does not affect the behavior of the program.
}
\label{slice-downside-example}
\end{figure}

\begin{figure}
\begin{lstlisting}
while true:
  x = input()
  if x != 0:
    break
  print("One over your number is: %d" % (1 / x))
\end{lstlisting}
\caption{
  In this example, the programmer attempts to handle invalid input by
  breaking out of the input loop. When computing the slice, our method
  would record a control dependency between the condition on line 3 and the
  print statement on line 5, placing the bad condition in the program
  slice.
}
\label{early-break}
\end{figure}

The goal of this feature, roughly equivalent in purpose to the type error
slice in Nate~\cite{learning-to-blame}, is to help eliminate terms that
cannot be the source of the crash. We compute a dynamic program slice
\cite{KOREL1988155}, the set of terms that contributed at runtime to the
observed exception, and turn this feature on for terms in the slice. We
follow the basic approach of \cite{KOREL1988155, KOREL1990187}, building a
graph of data and control dependencies. We then traverse the graph
backwards, starting at the execution step where the error occurred, to
collect the set of terms that the excepting line transitively depended on.
This excludes lines that could not have caused the exception, such as lines
that never ran, or lines that ran but had no connection to the step where
the exception happened.

A downside of using the dynamic slice, in the case of control dependencies, is
that it excludes terms whose presence does not affect whether the exception
happened, but which could have had an effect if the term was different. In
\figref{slice-downside-example}, the conditional statement is never
executed, so the removal of the conditional would not have any effect on
the program's behavior, and it would not be included in the program slice.
The dynamic slice misses the incorrect test condition because the mistake
itself caused the conditional statement and its body to become irrelevant.

To overcome this issues in the ``early return'' or ``early break'' case
(\figref{early-break}), we check if a break, return, or other statement for
escaping structured control flow is present inside of a conditional
statement. We then add dependencies in the execution trace's dependency graph
between the enclosing conditional and the statements that would have been skipped
by the break or return. While this heuristic is effective in practice, it
does not overcome all related problems: we thus treat dynamic slice
information as one of many features rather than as a hard constraint.
This boolean feature tracks whether or not the term is part of the final
dynamic slice.

\subsubsection{Crash Location (Dynamic)} We observe that the precise term that
causes Python to crash is, in fact, the one the user decides to change
much of the time. While this information is not sufficient on its own
(cf. the vanilla Python interpreter, which largely uses this feature), 
it can be a very helpful supplement. This boolean feature encodes whether
or not the term is exactly the source location of the uncaught exception.

\subsubsection{Exception Type (Dynamic)} The type of error thrown by the program
contains useful information its localization. In particular, a term may be
more or less suspicious depending on the ultimate uncaught exception. As an
example, a \lt{v / 0} term may be more suspicious if the uncaught exception
being considered is \lt{DivisionByZero}. We encode the exception type as
as a categorical feature. 
% to \emph{every} vector derived from that program.

\fixme{The original draft contained this text: ``For example, if asked how
likely a random `0` is of being a bug, you'd give a higher answer if you
knew that the program eventually crashes with a division by zero exception.''
This sort of rhetorical communication with the reader is \emph{too
informal for scientific writing}. I am not saying that I disagree with the
underlying argument. I am saying that you cannot really use the Socratic
method or the ``Appeal to Common Practice'' logical fallacy or the ``No
True Scotsman'' logical fallacy in scientific writing. If you wrote the
quoted text, read this FIXME carefully, ask me if you have any questions
about sloppy vs. formal arguments, and then delete this FIXME. --- Wes}

\subsubsection{Contextual features}

Term frequency vectors in general, and our bag of abstracted terms
representation by extension, do not typically include contextual
information such as ordering or relative positioning. As with many other
analyses, such as contextual operational semantics~\cite{FIXME}, we observe 
that the meaning of an expression may depend on its surrounding context.

As an example, consider the \lt{0} terms in both \lt{x = 0} and \lt{ x/ 0}.
Both instances of \lt{0} have the same syntactic form, size and type
(etc.). We may prefer to implicate the \lt{0} in \lt{x / 0} as suspicious,
especially for beginner-written programs, but cannot distinguish it without
surrounding contextual information. Data structures such as abstract syntax
trees and control flow graphics capture such contextual information, but
are not immediately applicable to machine learning.

We desire to encode such information while retaining the use of scalable,
accurate off-the-shelf machine learning algorithms that operate on feature
vectors. We thus propose to embed contextual information in a vector,
borrowing insights from standard approaches in machine learning. We
associated with term additional features that correspond to the features of
its parent and child nodes. For representational regularity, we always
model three children; terms without parents or children are given special
values (e.g., zero or NotApplicable) for those contextual features.

\subsection{Machine Learning Model}
\label{sec-model}
